---
title: An R Markdown document converted from "0.4 Data filtration.ipynb"
output: html_document
---

# Data Analysis Workshop - Data filtration

**Data source:**
<https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2022.1050574/full>


## Library Installation

```{r, eval=FALSE}
# Please run the following code to install all necessary libraries, 
# unless you have already installed them using conda recipe: conda_recipes/r_libraries.yaml.

# R Library Installation Script
# Install required packages for microbiome analysis

# Install BiocManager if not already installed (needed for Bioconductor packages)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install Bioconductor packages
BiocManager::install("phyloseq")
BiocManager::install("GUniFrac")

# Install CRAN packages
install.packages("vegan")
install.packages("ape")
install.packages("ggplot2")
install.packages("ggpubr")
install.packages("PMA")
install.packages("reshape2")
install.packages("tidyr")
install.packages("phytools")
install.packages("compositions")

# Verify installations by loading libraries
cat("\n=== Verifying installations ===\n")
libraries <- c("phyloseq", "vegan", "ape", "ggplot2", "ggpubr", 
               "PMA", "parallel", "reshape2", "tidyr", "GUniFrac", "phytools", "compositions")

for (lib in libraries) {
  if (require(lib, character.only = TRUE, quietly = TRUE)) {
    cat(sprintf("✓ %s loaded successfully\n", lib))
  } else {
    cat(sprintf("✗ %s failed to load\n", lib))
  }
}

cat("\nInstallation complete!\n")
```

## Load Libraries

```{r}
# Load necessary libraries for microbiome analysis
library(phyloseq)    # For microbiome data handling
library(vegan)       # For ecological diversity analysis
library(ape)         # For phylogenetic tree manipulation
library(ggplot2)     # For visualization
library(ggpubr)      # For publication-ready plots
library(PMA)         # For sparse CCA
library(parallel)    # For parallel computing
library(reshape2)    # For data reshaping
library(tidyr)       # For data tidying
library(GUniFrac)    # For UniFrac distance calculations
library(phytools)    # For phylogenetic tree tools
library(compositions) # For CLR transformation
```

```{r}
# Function to clean and standardize metadata
# This creates consistent sample IDs across microbiome and mycobiome datasets
clean_metadata <- function(metadata_preselected){
    metadata_modified <- metadata_preselected
    colnames(metadata_modified) <- c('age', 
                                     'sample_source',
                                     'sex',
                                     'subject_id')
    # Format age labels
    metadata_modified['age'] <- paste0('t', metadata_preselected[['host.age']], 'month')
    metadata_modified[metadata_preselected[['host.age']] == 6, 'age'] <- paste0(
        metadata_modified[metadata_preselected[['host.age']] == 6, 'age'], 's')
    
    # Simplify sample source names
    metadata_modified['sample_source'] <- gsub('human ', '', metadata_preselected[['host.body.product']])
    
    # Format subject IDs
    metadata_modified['subject_id'] <- paste0('fam#', metadata_modified[['subject_id']])
    
    # Create unique sample IDs
    metadata_modified['sample_id'] <- paste(metadata_modified[['subject_id']], 
                                        metadata_modified[['age']],
                                        metadata_modified[['sample_source']], sep='_')
    return(metadata_modified)
}
```

# Preprocessing 1

1. Select matching samples between microbiome and mycobiome datasets. <br>
2. Rename taxa with genera names.  <br>
3. Calculate relative abundances.  <br>

```{r}
# Define path to preprocessed phyloseq objects
data_path <- '../../data/PRJNA880162/processed/phyloseq'

# Create output directory for analysis results
dir.create("../../data/data_analysis", showWarnings = FALSE)
```

```{r}
# Load phyloseq objects containing microbiome (16S) and mycobiome (ITS) data
ps_microbiome <- readRDS('../../data/data_analysis/ps_microbiome.with_tree.rds')
ps_mycobiome  <- readRDS('../../data/data_analysis/ps_mycobiome.with_tree.rds')

# Display summary of loaded objects
ps_microbiome
ps_mycobiome
```

```{r}
# Preview OTU table structure
otu_table_preview <- data.frame(head(otu_table(ps_microbiome)))

# Preview taxonomy table structure
tax_table_preview <- data.frame(head(tax_table(ps_microbiome)))
```

```{r}
# Check for and remove samples with zero total abundance
nrow(otu_table(ps_microbiome)[rowSums(otu_table(ps_microbiome)) == 0, ])
nrow(otu_table(ps_mycobiome)[rowSums(otu_table(ps_mycobiome)) == 0, ])

# Prune samples with zero abundance
ps_microbiome <- prune_samples(rowSums(otu_table(ps_microbiome)) > 0, ps_microbiome)
ps_mycobiome <- prune_samples(rowSums(otu_table(ps_mycobiome)) > 0, ps_mycobiome)

ps_microbiome
ps_mycobiome
```

## Convert to Relative Abundance

```{r}
# Transform counts to relative abundances (proportions)
# This normalizes for different sequencing depths across samples
ps_microbiome <- transform_sample_counts(ps_microbiome, function(x) {x / sum(x)})
ps_mycobiome <- transform_sample_counts(ps_mycobiome, function(x) {x / sum(x)})
```

```{r}
# Verify the transformation (values should sum to 1 for each sample)
otu_table_preview <- data.frame(head(otu_table(ps_microbiome)))
```

## Subset Same Samples

```{r}
# Verify the transformation (values should sum to 1 for each sample)
metadata_table_preview <- data.frame(head(sample_data(ps_microbiome)))[, c('host.age', 
                                                                           'host.body.product', 
                                                                           'host.sex', 
                                                                           'host.subject.id')]
```

```{r}
# Clean microbiome metadata
metadata_microbiome <- data.frame(sample_data(ps_microbiome))
metadata_microbiome <- metadata_microbiome[, c('host.age', 
                                               'host.body.product', 
                                               'host.sex', 
                                               'host.subject.id')]
metadata_microbiome <- clean_metadata(metadata_microbiome)
sample_data(ps_microbiome) <- metadata_microbiome
sample_names(ps_microbiome) <- metadata_microbiome[['sample_id']]
```

```{r}
# Verify the transformation (values should sum to 1 for each sample)
metadata_table_preview <- data.frame(head(sample_data(ps_microbiome)))
```

```{r}
# Clean mycobiome metadata
metadata_mycobiome <- data.frame(sample_data(ps_mycobiome))
metadata_mycobiome <- metadata_mycobiome[, c('host.age', 
                                             'host.body.product', 
                                             'host.sex', 
                                             'host.subject.id')]
metadata_mycobiome <- clean_metadata(metadata_mycobiome)
sample_data(ps_mycobiome) <- metadata_mycobiome
sample_names(ps_mycobiome) <- metadata_mycobiome[['sample_id']]
```

```{r}
# Find samples present in both datasets
sample_ids_microbiome <- metadata_microbiome[['sample_id']]
sample_ids_mycobiome <- metadata_mycobiome[['sample_id']]
sample_ids <- intersect(sample_ids_microbiome, sample_ids_mycobiome)
```

```{r}
# Subset to common samples and ensure same order
ps_microbiome <- prune_samples(sample_ids, ps_microbiome)
ps_mycobiome <- prune_samples(sample_ids, ps_mycobiome)

# Ensure samples are in the same order in both datasets
otu_table(ps_microbiome) <- otu_table(ps_microbiome)[sample_ids, ]
otu_table(ps_mycobiome) <- otu_table(ps_mycobiome)[sample_ids, ]

ps_microbiome
ps_mycobiome
```

```{r}
# Save processed phyloseq objects for later use
saveRDS(ps_microbiome, '../../data/data_analysis/ps_microbiome.filtered.rds')
saveRDS(ps_mycobiome, '../../data/data_analysis/ps_mycobiome.filtered.rds')
```

# Preprocessing 2. Agglomerated by genus

1. Select matching samples between microbiome and mycobiome datasets. <br>
2. Agglomerate by Genus.  <br>
3. Rename taxa with genera names.  <br>
4. Calculate relative abundances.  <br>

```{r}
# Define path to preprocessed phyloseq objects
data_path <- '../../data/PRJNA880162/processed/phyloseq'

# Create output directory for analysis results
dir.create("../../data/data_analysis", showWarnings = FALSE)
```

```{r}
# Load phyloseq objects containing microbiome (16S) and mycobiome (ITS) data
ps_microbiome <- readRDS('../../data/data_analysis/ps_microbiome.with_tree.rds')
ps_mycobiome  <- readRDS('../../data/data_analysis/ps_mycobiome.with_tree.rds')

# Display summary of loaded objects
ps_microbiome
ps_mycobiome
```

```{r}
# Preview OTU table structure
otu_table_preview <- data.frame(head(otu_table(ps_microbiome)))

# Preview taxonomy table structure
tax_table_preview <- data.frame(head(tax_table(ps_microbiome)))
```

## Agglomerate by Genus

```{r}
# Agglomerate taxa at the genus level
# This combines all OTUs belonging to the same genus
ps_microbiome <- tax_glom(ps_microbiome, 'Genus')
ps_mycobiome <- tax_glom(ps_mycobiome, 'Genus')

ps_microbiome
ps_mycobiome
```

```{r}
# Rename taxa using genus names for easier interpretation
taxa_names(ps_microbiome) <- tax_table(ps_microbiome)[, 'Genus']
taxa_names(ps_mycobiome) <- tax_table(ps_mycobiome)[, 'Genus']
```

```{r}
# Verify the renaming
otu_table_preview <- data.frame(head(otu_table(ps_microbiome)))
```

```{r}
# Check for and remove samples with zero total abundance
nrow(otu_table(ps_microbiome)[rowSums(otu_table(ps_microbiome)) == 0, ])
nrow(otu_table(ps_mycobiome)[rowSums(otu_table(ps_mycobiome)) == 0, ])

# Prune samples with zero abundance
ps_microbiome <- prune_samples(rowSums(otu_table(ps_microbiome)) > 0, ps_microbiome)
ps_mycobiome <- prune_samples(rowSums(otu_table(ps_mycobiome)) > 0, ps_mycobiome)

ps_microbiome
ps_mycobiome
```

## Convert to Relative Abundance

```{r}
# Transform counts to relative abundances (proportions)
# This normalizes for different sequencing depths across samples
ps_microbiome <- transform_sample_counts(ps_microbiome, function(x) {x / sum(x)})
ps_mycobiome <- transform_sample_counts(ps_mycobiome, function(x) {x / sum(x)})
```

```{r}
# Verify the transformation (values should sum to 1 for each sample)
otu_table_preview <- data.frame(head(otu_table(ps_microbiome)))
```

## Subset Same Samples

```{r}
# Verify the transformation (values should sum to 1 for each sample)
metadata_table_preview <- data.frame(head(sample_data(ps_microbiome)))[, c('host.age', 
                                                                           'host.body.product', 
                                                                           'host.sex', 
                                                                           'host.subject.id')]
```

```{r}
# Clean microbiome metadata
metadata_microbiome <- data.frame(sample_data(ps_microbiome))
metadata_microbiome <- metadata_microbiome[, c('host.age', 
                                               'host.body.product', 
                                               'host.sex', 
                                               'host.subject.id')]
metadata_microbiome <- clean_metadata(metadata_microbiome)
sample_data(ps_microbiome) <- metadata_microbiome
sample_names(ps_microbiome) <- metadata_microbiome[['sample_id']]
```

```{r}
# Verify the transformation (values should sum to 1 for each sample)
metadata_table_preview <- data.frame(head(sample_data(ps_microbiome)))
```

```{r}
# Clean mycobiome metadata
metadata_mycobiome <- data.frame(sample_data(ps_mycobiome))
metadata_mycobiome <- metadata_mycobiome[, c('host.age', 
                                             'host.body.product', 
                                             'host.sex', 
                                             'host.subject.id')]
metadata_mycobiome <- clean_metadata(metadata_mycobiome)
sample_data(ps_mycobiome) <- metadata_mycobiome
sample_names(ps_mycobiome) <- metadata_mycobiome[['sample_id']]
```

```{r}
# Find samples present in both datasets
sample_ids_microbiome <- metadata_microbiome[['sample_id']]
sample_ids_mycobiome <- metadata_mycobiome[['sample_id']]
sample_ids <- intersect(sample_ids_microbiome, sample_ids_mycobiome)
```

```{r}
# Subset to common samples and ensure same order
ps_microbiome <- prune_samples(sample_ids, ps_microbiome)
ps_mycobiome <- prune_samples(sample_ids, ps_mycobiome)

# Ensure samples are in the same order in both datasets
otu_table(ps_microbiome) <- otu_table(ps_microbiome)[sample_ids, ]
otu_table(ps_mycobiome) <- otu_table(ps_mycobiome)[sample_ids, ]

ps_microbiome
ps_mycobiome
```

```{r}
# Save processed phyloseq objects for later use
saveRDS(ps_microbiome, '../../data/data_analysis/ps_microbiome.genus_agglomerated.rds')
saveRDS(ps_mycobiome, '../../data/data_analysis/ps_mycobiome.genus_agglomerated.rds')
```

